### 2024-12-04 teaching/math240f24/draft.tex
Last lecture

\medskip\boldlabel\datestamp{26}{} Recurrences with constant coefficients.
We now turn to the case where the $f_i$ are all constant functions. This means
there are constants $c_1, c_2, \ldots, c_k$ such that $f_1(n) = c_1$ for all $n$,
$f_2(n) = c_2$ for all $n$, and so on.
In light of Theorem~{\thmnonhomo}, we can
restrict our attention to homogeneous recurrences, since we can solve non-homogeneous
recurrences by solving their associated homogeneous recurrences and then applying the theorem.
Homogeneous recurrences with constant coefficients look like
$$a_n = c_1 a_{n-1} + c_2 a_{n-2} + \cdots + c_k a_{n-k},$$
for some degree $k\ge 1$. We define the {\it characteristic equation} of such a recurrence to be the
equation
$$x^k = c_1 x^{k-1} + c_2 x^{k-2} + \cdots + c_k,$$
and the {\it characteristic polynomial} is the polynomial obtained by subtracting
the right-hand side from the left-hand side:
$$x^k - c_1 x^{k-1} + c_2 x^{k-2} + \cdots + c_k$$
It turns out that this
polynomial entirely determines the general solution to the recurrence. We shall state
and prove this theorem in the special case that $k=2$ and the two roots of the characteristic
polynomial are distinct.

\edef\thmdegtwodistinct{\the\thmcount}
\proclaim Theorem \advthm. Consider the recurrence given by
$$a_n = c_1 a_{n-1} + c_2 a_{n-2}.$$
If the characteristic polynomial of this recurrence has two distinct roots $r_1$ and $r_2$, then
the general solution of the recurrence is
$$a_n = \alpha_1 {r_1}^n + \alpha_2 {r_2}^n.$$
Supposing we know the initial conditions $a_0$ and $a_1$, we can solve for
$\alpha_1$ and $\alpha_2$ via the identities
$$\alpha_1 = {a_0 r_2 - a_1\over r_2 - r_1}
\qquad\hbox{and}\qquad
\alpha_2 = {a_1 - a_0 r_1\over r_2-r_1}.$$

@@

The case where the characteristic polynomial has a repeated root is not covered by
Theorem~{\thmdegtwodistinct}. Instead, we have the following theorem, which we will
not have time to prove in this course (though the proof is by no means difficult).

\proclaim Theorem \advthm. Consider the recurrence given by
$$a_n = c_1 a_{n-1} + c_2 a_{n-2}.$$
If the characteristic polynomial of this recurrence has a repeated root $r$,
then the general solution of the recurrence is
$$a_n = \alpha_1 r^n + \alpha_2 nr^n.\noskipslug$$

Theorem~{\thmdegtwodistinct} can also be generalised to higher degrees $k$. Its proof is
morally the same as the one we gave for the case $k=2$.

@@

>>>>

\medskip\boldlabel\datestamp{03}{xii} Recurrences with constant coefficients.
We now turn to the case where the $f_i$ are all constant functions. This means
there are constants $c_1, c_2, \ldots, c_k$ such that $f_1(n) = c_1$ for all $n$,
$f_2(n) = c_2$ for all $n$, and so on.
In light of Theorem~{\thmnonhomo}, we can
restrict our attention to homogeneous recurrences, since we can solve non-homogeneous
recurrences by solving their associated homogeneous recurrences and then applying the theorem.
Homogeneous recurrences with constant coefficients look like
$$a_n = c_1 a_{n-1} + c_2 a_{n-2} + \cdots + c_k a_{n-k},$$
for some degree $k\ge 1$. We define the {\it characteristic equation} of such a recurrence to be the
equation
$$x^k = c_1 x^{k-1} + c_2 x^{k-2} + \cdots + c_k,$$
and the {\it characteristic polynomial} is the polynomial obtained by subtracting
the right-hand side from the left-hand side:
$$x^k - c_1 x^{k-1} + c_2 x^{k-2} + \cdots + c_k$$
It turns out that this
polynomial entirely determines the general solution to the recurrence. We shall state
and prove this theorem in the special case that $k=2$ and the two roots of the characteristic
polynomial are distinct.

\edef\thmdegtwodistinct{\the\thmcount}
\proclaim Theorem \advthm. Consider the recurrence given by
$$a_n = c_1 a_{n-1} + c_2 a_{n-2}.$$
If the characteristic polynomial of this recurrence has two distinct roots $r_1$ and $r_2$, then
the general solution of the recurrence is
$$a_n = \alpha_1 {r_1}^n + \alpha_2 {r_2}^n.$$
Supposing we know the initial conditions $a_0$ and $a_1$, we have the identities
$$\alpha_1 = {a_0 r_2 - a_1\over r_2 - r_1}
\qquad\hbox{and}\qquad
\alpha_2 = {a_1 - a_0 r_1\over r_2-r_1}.$$

@@

Theorem~{\thmdegtwodistinct} can be generalised to higher degrees $k$. Its proof is
morally the same as the one we gave for the case $k=2$.

@@

Returning to the setting $k=2$,
the case where the characteristic polynomial has a repeated root is not covered by
Theorem~{\thmdegtwodistinct}. Instead, we have the following theorem.

\edef\thmdegtworepeated{\the\thmcount}
\proclaim Theorem \advthm. Consider the homogeneous recurrence
$$a_n = c_1 a_{n-1} + c_2 a_{n-2}.$$
If the characteristic polynomial of this recurrence has a repeated root $r$,
then the general solution of the recurrence is
$$a_n = \alpha_1 r^n + \alpha_2 nr^n.$$
Supposing we know the initial conditions $a_0$ and $a_1$, we have the identities
$$ \alpha_1 = a_0\qquad\hbox{and}\qquad \alpha_2 = {a_1 - a_0 r\over r}.$$

\proof By the same logic as in the proof of Theorem~{\thmdegtwodistinct}, we know that
$r^n$ is a solution to the recurrence. We show that $b_n = nr^n$ is also a solution.
Since $r$ is a repeated root to the characteristic polynomial $x^2 - c_1x - c_2$, we have
$$x^2 - c_1 x - c_2 = (x-r)^2 = x^2 - 2rx + r^2,$$
so by comparing coefficients we conclude that $c_1 = 2r$ and $c_2 = -r^2$.
Now by expanding
$$\eqalign{
c_1 b_{n-1} + c_2 b_{n-2} &= c_2(n-1)r^{n-1} + c_2 (n-2) r^{n-2} \cr
&= r^{n-2} \bigl( c_1 (n-1) r + c_2(n-2)\bigr) \cr
&= r^{n-2} (c_1 n r - c_1 r + c_2 n - 2c_2 ) \cr
&= r^{n-2} (2r^2 n - 2r^2 -r^2 n + 2r^2 ) \cr
&= nr^n \cr
&= b_n \cr
}$$
we see that $b_n = nr^n$ satisfies the recurrence as well. By similar reasoning
as in the proof of Theorem~{\thmdegtwodistinct}, any linear combination of the two solutions
$(r^n)$ and $(nr^n)$ is also a solution.

Now we prove the further claim. Suppose we know $a_0$ and $a_1$. Then
$$a_0 = \alpha_1 r^0 + \alpha_2 0r^0 = \alpha_1,$$
and
$$ a_1 = \alpha_1 r^1 + \alpha_2 r^1 = a_0 r + \alpha_2 r,$$
and hence
$$ \alpha_1 = a_0\qquad\hbox{and}\qquad \alpha_2 = {a_1 - a_0 r\over r}.\noskipslug$$

Let's illustrate this theorem with a final example. We solve the recurrence
$$a_n = 4 a_{n-1} - 4 a_{n-2} + n,$$
with the initial conditions $a_0 = 1$ and $a_1 = 3$.
First we solve the homogeneous recurrence
$$ h_n = 4h_{n-1} + 4 h_{n-2}.$$
The characteristic polynomial is $x^2 - 4x + 4 = (x-2)^2$, so by the above theorem,
$$h_n = \alpha_1 2^n + \alpha_2 n 2^n.$$
Now we need a particular solution $p_n$. We have $g(n) = n$, which is a linear function
in $n$, so we guess that $p_n = cn+d$ for some real numbers $c$ and $d$. Then
from the identity $p_n = 4p_{n-1} - 4p_{n-2} + n$, we have
$$\eqalign{
cn+d &=  4\bigl( c(n-1) + d\bigr) - 4 \bigl( c(n-2)+d\bigr) + n \cr
0 &= -cn - d + 4cn - 4c + 4d - 4cn + 8c - 4d + n \cr
0 &= (1-c) n + (4c-d).
}$$
So $1-c = 0$ and $4c-d = 0$, whence $c = 1$ and $d=4$. So $p_n = n+4$.
This gives us the general solution
$$ a_n = \alpha_1 2^n + \alpha_2 n 2^n + n + 4.$$
We cannot use Theorem~{\thmdegtworepeated} to get $\alpha_1$ and $\alpha_2$, since
that theorem concerned homogeneous recurrences. Instead we directly compute
$$ 1 = a_0 = \alpha_1 2^0 + \beta\cdot 0 + 0 + 4 = \alpha_1 + 4,$$
and derive $\alpha_1 = -3$, then find that
$$3 = a_1 = \alpha_1 2 + \alpha_1 \cdot 2 + 1 + 4
= -6 + 2\alpha_2 + 4.$$
Hence $4=2\alpha_2$ and $\alpha_2 = 2$, which gives us a final solution of
$$a_n = 3\cdot 2^n + 2n\cdot 2^n + n + 4$$
to our recurrence.
